Avoiding global deforestation by taxing land in agricultural production: The implications for global markets

### 2024-03-09 paper.Rmd
Update paper.Rmd

In 2020, there were 4.1 billion hectares of forestland globally covering roughly 31.4% of the land [@fao2023]. These forests play an important role in maintaining and growing the land-based carbon sink by sequestering about 30% of GHG emissions currently [@roe2019]. This is because of forests proven ability to sequester carbon not only in the stems of trees [@davis2022] but in their soils as well [@guo2021]. With increasingly negative climate impacts, the pressure to enhance the terrestrial sink is growing, and one of the tools to achieve this is afforestation. As such, there is growing attention being placed globally to afforestation efforts, which is highlighted by the UN's declaration of the years from 2021 through 2030 as the UN Decade of Ecosystem Restoration [@sewell2020]. Many of the Nationally Determined Contributions (NDCs) countries have submitted, which detail their plans to keep warming under the 2C threshold, also rely to a significant degree on afforestation projects [@roe2019].

The global gains that can be realized from afforestation, however, are uncertain, with broad estimates of cumulative sequestration ranging from 42 gigatonnes to 700 gigatonnes by the end of the century [@veldman2019; @bastin2019; @humpenoder2014; @sathaye2006; @doelman2020]. In addition, the amount of land conversion that would be required also varies greatly, with estimates ranging from 0.3 billion hectares to 2.8 billion hectares [@veldman2019; @bastin2019; @humpenoder2014; @doelman2020]. In many cases, these results suggest massive losses in the amount of pasture and cropland that would be available. This puts into question the feasibility of attaining these results, as almost 90 percent of deforestation worldwide is undertaken to expand the amount of agricultural land [@fao2021]. There are many reasons for this. For example, as nations develop and disposable incomes rise, consumers often increase meat consumption. Population growth is also a driver. Currently, the United Nations' medium variant population projections suggest that between 2023 and 2033 the worlds' population will grow by 8.8 percent, a gain of 707.1 million people [@un2024]. And where population growth is especially high, which is the case at present throughout Africa (2023 to 2033 population growth: 25.0%: 360.6 million), there is likely to be greater conversion of forestland into farmland. Still, some research has suggested that substantive gains in the carbon sink (+23.8PgCo2e/yr) can be made without sacrificing agricultural land [@griscom2017].

If such efforts are to be made, it is important to understand where to target them, as afforestation does not provide equal carbon benefits in every region. For example, afforestation in boreal zones provides the smallest net benefits, as the carbon sequestration benefits of conifer forests are mitigated by the reduction in albedo in winter [@mykleby2017; @li2015]. Coniferous forests have also been shown to increase soil organic carbon less than broadleaf forests [@guo2021]. Conversely, tropical forests have been linked with a strong positive result from afforestation, as the carbon sequestration benefit is complemented by a cooling impact due to changes in both albedo and evapotranspiration, and with temperate forests, the impact of afforestation is also shown to be positive, albeit less strongly so due to a winter warming effect [@li2015].

South America and Sub-Saharan Africa are thus two regions that show much potential for afforestation. It is estimated that these two regions hold at least 50% of the potential for global gains while regions such as Northern Africa and the Middle East show little promise as forest growth rates are quite low [@doelman2020]. Thus, it is promising that the African Forest Landscape Restoration Initiative (AFR100), which aims to afforest 100 million hectares by 2030, has roughly 70% of its commitments coming from Sub-Saharan nations [@auda2023]. There are other nations that have a big potential for gains from afforestation, but they face challenges. India, for example, has a great potential to increase its carbon sequestration through afforestation, but its massive population and food security worries limit its ability to consider such actions. The United Nations suggests that India's population will grow less rapidly than the world average between 2023 and 2033 (+8.5%), but due to its sizable base, the increase will be substantial nonetheless: 120.5 million. India's NDC does not entertain the idea of agricultural land being re-purposed and instead primarily considers targeting land that may not be well suited for afforestation [@amjath2019]. Moreover, as land-use decisions in one country also tend to have impacts that spillover into many others [@meyfroidt2010; @lambin2011], analysis needs to be conducted at a global scale.

Several notable modeling attempts have been made in this direction. @steinbuks2012 developed their global partial equilibrium FABLE model (Forestry, Agriculture, Biofuels, Land Use, and Environment) to identify the optimal land use choices given the competing demands of meeting greenhouse gas targets and satisfying demand for food, bioenergy, forest products, and ecosystem services. While FABLE is a powerful tool for examining the optimal trajectory of various land uses under specific demand assumptions, it is focused on specific sectors of the economy and neglects general equilibrium and household welfare effects [@steinbuks2012; @hertel2017]. Other efforts have used the Global Trade Analysis Project (GTAP) computable general equilibrium model to address such questions, but its standard model does not well account for land-use changes. This prompted the creation of the Land Use and Land Cover Database within the GTAP framework, which incorporates forestry remote-sensing products [@baldos2012]. The GTAP Agro-Ecological Zone (GTAP-AEZ) model further strengthened the GTAP model's ability to analyze land use changes, with the use of spatially explicit global land use data and through the incorporation of intra- and inter-regional land and land-based greenhouse emissions heterogeneity [@hertel2008]. While the GTAP-AEZ model benefits from the inclusion of competition among different crops, grazing, and forest-based uses, it has limitations, primarily due to its infrequent updates. The KLUM@GTAP framework integrated the Kleines Land Use Model (KLUM) with an extended yet static version of GTAP called GTAP-EFL to evaluate the impact of climate change on cropland allocation. KLUM is a global agricultural land-use model that connects the economy to global crop allocation to maximize producer returns under specific risk assumptions, and GTAP-EFL separates energy factors from intermediate inputs and incorporates them into capital, while also considering CO2 emissions. KLUM@GTAP substitutes the land allocation mechanism within GTAP-EFL by utilizing regionally aggregated area changes in cropland determined by KLUM to update cropland shares in GTAP-EFL [@ronneberger2009].

More recent modeling attempts of land-use change have begun to move away from comparative static analysis though. For example, the dynamic GTAP model, supported by the GTAP-AEZ database, considers land market effects, which were identified as significant in driving results by @stevenson2013. This model has been utilized to evaluate the impact of crop intensification on land use. It, however, has been critiqued, as global aggregates may mask localized shifts that can have implications for ecologically significant areas [@byerlee2014]. In addition, in the standard GTAP model, it is difficult to model a sector individually. There are 68 market sectors covered in the latest version of the GTAP database. The GTAP model's general equilibrium nature means that all these markets are cleared with market-clearing prices. As the model applies nested CES production functions to all sectors and differentiates them by parameters only, to model one sector, in this case forestry, differently from the rest would require a major redesign of the model. While not impossible, this project will attempt to get around these concerns and assess the impact of various likely land-use scenarios by melding the dynamic GTAP model with the Global Timber Model (GTM), which is a dynamic optimization model that has been specifically designed to analyze the relationship among land rents, forested land cover, and carbon sequestration through forests and thus is better able to examine the impact of afforestation or avoided deforestation on the carbon sink [@sohngen2001]. The latest versions of both models were used. For the GTAP model, this meant using data which has 2017 as its base year. This data was then updated to the base year (2023) using data from the Centre for Prospective Studies and International Information (CEPII) [@CEPII2024] to account for the projected changes in gross domestic product (GDP), population, and productivity for each country in the world. Similar steps were taken to update and recalibrate the GTM model. As most forest inventory data are not updated annually, the GTM model updates data for regions with inventory information approximately every five years, and it updates data for other regions using the Global Forest Resource Assessment [@FAO2020].

@@

These reductions in the amount of agricultural land would then impact the quantity of agricultural goods that could be produced. For example, the 15.0 percent decrease in agricultural land in East Asia and the 9.0 percent decrease in agricultural land in Southeast Asia would result in reductions in agricultural production of 5.0 percent and 1.7 percent, respectively. Surprisingly, Brazil and the United States, despite their decreases in available agricultural land compared to the baseline 2033 scenario, are estimated to see increased agricultural output of 0.6 percent and 1.0 percent, respectively. This happens when the sharp output decline in other regions prompts increased intensity of production in the United States and Brazil.

>>>>

In 2020, there were 4.1 billion hectares of forestland globally covering roughly 31.4% of the land [@fao2023]. These forests play an important role in maintaining and growing the land-based carbon sink by sequestering about 30% of GHG emissions currently [@roe2019]. This is because of forests proven ability to sequester carbon not only in the stems of trees [@davis2022] but in their soils as well [@guo2021]. With increasingly negative climate impacts, there is increased attention being placed on maintaining these carbon sinks, but this requires, at minimum, avoiding further deforestation. However, from 2015 to 2020, roughly 10 million hectares of forestland was converted annually to other land uses [@]. There are many reasons for this. For example, as nations develop and disposable incomes rise, consumers often increase meat consumption. Population growth is also a driver. Currently, the United Nations' medium variant population projections suggest that between 2023 and 2033 the world's population will grow by 8.8 percent, a gain of 707.1 million people [@un2024], and where population growth is especially high, which is the case at present throughout Africa (2023 to 2033 population growth: 25.0%: 360.6 million), there is likely to be greater conversion of forestland into farmland. This puts into question whether avoiding further deforestation is attainable.

Moreover, if efforts to end deforestation are to be made, it is important to understand where to target them, as forests do not provide equal carbon benefits in every region. For example, conifer forestland in boreal zones provides the smallest net benefits, as gains in carbon sequestration are mitigated by reductions in albedo in winter [@mykleby2017; @li2015]. Coniferous forests have also been shown to increase soil organic carbon less than broadleaf forests [@guo2021]. Conversely, tropical forests have been linked with a strong positive results, as the carbon sequestration benefit is complemented by a cooling impact due to changes in both albedo and evapotranspiration, and with temperate forests, the impact is also shown to be positive, albeit less strongly so due to a winter warming effect [@li2015]. As land-use decisions in one country also tend to have impacts that spillover into many others [@meyfroidt2010; @lambin2011], analysis needs to be conducted at a global scale.

Several notable modeling attempts have been made in this direction. @steinbuks2012 developed their global partial equilibrium FABLE model (Forestry, Agriculture, Biofuels, Land Use, and Environment) to identify the optimal land use choices given the competing demands of meeting greenhouse gas targets and satisfying demand for food, bioenergy, forest products, and ecosystem services. While FABLE is a powerful tool for examining the optimal trajectory of various land uses under specific demand assumptions, it is focused on specific sectors of the economy and neglects general equilibrium and household welfare effects [@steinbuks2012; @hertel2017]. Other efforts have used the Global Trade Analysis Project (GTAP) computable general equilibrium model to address such questions, but its standard model does not well account for land-use changes. This prompted the creation of the Land Use and Land Cover Database within the GTAP framework, which incorporates forestry remote-sensing products [@baldos2012]. The GTAP Agro-Ecological Zone (GTAP-AEZ) model further strengthened the GTAP model's ability to analyze land use changes, with the use of spatially explicit global land use data and through the incorporation of intra- and inter-regional land and land-based greenhouse emissions heterogeneity [@hertel2008]. While the GTAP-AEZ model benefits from the inclusion of competition among different crops, grazing, and forest-based uses, it has limitations, primarily due to its infrequent updates.

More recent modeling attempts of land-use change have begun to move away from comparative static analysis though. For example, the dynamic GTAP model, supported by the GTAP-AEZ database, considers land market effects, which were identified as significant in driving results by @stevenson2013. This model has been utilized to evaluate the impact of crop intensification on land use. It, however, has been critiqued, as global aggregates may mask localized shifts that can have implications for ecologically significant areas [@byerlee2014]. In addition, in the standard GTAP model, it is difficult to model a sector individually. There are 68 market sectors covered in the latest version of the GTAP database. The GTAP model's general equilibrium nature means that all these markets are cleared with market-clearing prices. As the model applies nested CES production functions to all sectors and differentiates them by parameters only, to model one sector, in this case forestry, differently from the rest would require a major redesign of the model.

This project gets around these concerns and assesses the impact of various likely land-use scenarios by melding the dynamic GTAP model with the Global Timber Model (GTM), which is a dynamic optimization model that has been specifically designed to analyze the relationship among land rents, forested land cover, and carbon sequestration through forests and thus is better able to examine the impact of afforestation or avoided deforestation on the carbon sink [@sohngen2001]. The latest versions of both models were used. For the GTAP model, this meant using data which has 2017 as its base year. This data was then updated to the base year (2023) using data from the Centre for Prospective Studies and International Information (CEPII) [@CEPII2024] to account for the projected changes in gross domestic product (GDP), population, and productivity for each country in the world. Similar steps were taken to update and recalibrate the GTM model. As most forest inventory data are not updated annually, the GTM model updates data for regions with inventory information approximately every five years, and it updates data for other regions using the Global Forest Resource Assessment [@FAO2020].

@@

These reductions in the amount of agricultural land would then impact the quantity of agricultural goods that could be produced. For example, the 15.0 percent decrease in agricultural land in East Asia and the 9.0 percent decrease in agricultural land in Southeast Asia would result in reductions in agricultural production of 5.0 percent and 1.7 percent, respectively. Surprisingly, Brazil and the United States, despite their decreases in available agricultural land compared to the baseline 2033 scenario, are estimated to see increased agricultural output of 0.6 percent and 1.0 percent, respectively. This is the result of the sharp output declines in other regions prompting increased intensity of production in the United States and Brazil.
